ANNO: The Anthropological Notation Ontology

### 2023-03-30 paper.tex
Improve related work on nomenclatures.

\section{Related Work}
The Foundational Model of Anatomy (FMA) ontology \cite{fma} describes concepts of anatomy.
\url{https://bioportal.bioontology.org/ontologies/FMA}

\section{Ontology Development}
%TODO: lohnt sich nicht als separate Sektion, woanders mit unterbringen
% @Alex ist das CSV->OWL Script irgendwo publiziert, kann man das zitieren?
The domain experts are provided a spreadsheet-based input template by the ontologists,
which is transformed to OWL using a custom script.
% ist das richtig übersetzt?
Sources are categorized into core, root, and occurrence sources.

\section{Ontology Modelling}
While there are many textual sources of anthropological systematization, they do not agree in all aspects and there is not a single, formally described, standard.
ANNO links to concepts of the FMA when they exist, but structures them in a different way.
The ontology consists of the following core modules:
%For the categories, there were the following spreadsheets: bones, anatomical landmarks, osteometric landmarks, measured distances, function (separately for discriminant and regress functions), anatomical position descriptions, sources used, and osteological terminology.
\subsection{Bones}

>>>>

\section{Related Work}
\subsection{Anatomical Nomenclatures}
There are various nomenclatures for the description of anatomical landmarks.
These allow a uniform classification of anatomical features and structures:

The \emph{Terminologia anatomica}~\citep{ta2} (TA) is a hierarchy of anatomical landmarks of the entire human body.
From its history, the Latin and Greek designations have emerged to allow standardization.
It is advantageous that these languages no longer show progress and are increasingly losing political power. 
In TA, the order of structures follows the natural anatomy.
An anatomical landmark is identified by an individual identification number, the anatomical term, and the English term.
The identification number is critical because certain landmarks are redundant.
For example, there is the \emph{processus zygomaticus} at the \emph{os frontale} and the \emph{os temporale}.
Synonyms, like \emph{zygomatic bone} and \emph{malar bone}, are also given.
Terms included in the \emph{other} category given include eponyms, which are concepts named after people, places or things.
For example, \emph{ossa digitorum pedis} (engl. \emph{phalanges of the foot}) are the small bones of the toes and are also denoted as \emph{ossa phalangea pedis} which are named after the Greek word \emph{phalanx} for a dense rectangular infantry formation.
% chapter 2: Ossa (Bones)

The \emph{Foundational Model of Anatomy} (FMA) ontology~\citep{fma} represents the physical organization of human anatomy.
%It allows the knowledge it contains to be represented in a human-understandable and machine-interpretable way.
However, the Latin and Greek designations are not taken into account, due to hypernyms 90\,\% of the terms are re-designated and only 1\,\% contain definitions.
This has a negative impact on the comprehensibility for users.

Furthermore, in both nomenclatures there is an inconsistent use of singular and plural forms, gaps in lateralized landmarks and numbered series of certain bones (e.g. costae).
In anatomy, it is also important to distinguish between the medical and anthropological viewpoints.
For example, relevant anthropological aspects such as the degree of wear of bones are not considered.
Furthermore, the articular surfaces of the individual bones are particularly important in anthropology.
Not all articular surfaces (e.g., on the ossa metatarsalia) are defined in the nomenclatures, resulting in further gaps.
In some cases, contradictions are also found in the definitions of individual landmarks.
For example, the corpus ossis pubis is placed elsewhere in some sources than in others.
Another problem is that the individual services do not synchronize and do not offer enough synonyms. % which services? 
In addition, the anatomical landmarks are not visually represented and are only selectively labeled in known anatomy atlases.
However, since landmarks are three-dimensional structures, it is essential to provide a marker over the entire landmark and delineate it accurately.
The delineations must also be clearly described textually.
However, there are no standardized definitions in anatomy so far and also in the literature the information is sparse.
For example, in some sources only punctual labels are presented on a figure.
A similar situation exists with osteometric landmarks.
Standardized ones are found only on the cranium and mandible.
However, since there are measurement distances on all bones, the respective start and end points must also be defined.
These are stated differently in the literature and have a high number of synonyms.

For these reasons, it is necessary to develop uniform, textual, visual and detailed descriptive definitions for the individual landmarks, as well as to represent them in a new and clear way in an ontology.
The ontology should be easy to understand for the user.

%\url{https://bioportal.bioontology.org/ontologies/FMA}

\subsection{\aw}

\section{Ontology Modelling}
% @Alex ist das CSV->OWL Script irgendwo publiziert, kann man das zitieren?
The domain experts are provided a spreadsheet-based input template by the ontologists,
which is transformed to OWL using a custom script.
% ist das richtig übersetzt?
Sources are categorized into core, root, and occurrence sources.

While there are many textual sources of anthropological systematization, they do not agree in all aspects and there is not a single, formally described, standard.
ANNO links to concepts of the FMA when they exist, but structures them in a different way.
The ontology consists of the following core modules:
%For the categories, there were the following spreadsheets: bones, anatomical landmarks, osteometric landmarks, measured distances, function (separately for discriminant and regress functions), anatomical position descriptions, sources used, and osteological terminology.
\subsection{Bones}
